### 2015-05-25 paper/manuscript/plos_template.tex
organized directory

% Template for PLoS
% Version 2.0 July 2014
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% Be advised that this is merely a template 
% designed to facilitate accurate translation of manuscript content 
% into our production files. 
%
% This template contains extensive comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template 
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication and enters production, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% DO NOT ADD EXTRA PACKAGES TO THIS TEMPLATE unless absolutely necessary.
% Packages included in this template are intentionally
% limited and basic in order to reduce the possibility
% of issues during our production process.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For sideways tables, use the {rotating} package and use \begin{sidewaystable} instead of \begin{table} in the appropriate section. PLOS guidelines do not accomodate sideways figures.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment, and for any superscripts or subscripts also include the base number/text. For example, use $mathrm{mm}^2$ instead of mm$^2$ (do not use \textsuperscript command).
%
% DO NOT USE the \rm command to render mathmode characters in roman font, instead use $\mathrm{}$
% For bolding characters in mathmode, please use $\mathbf{}$ 
%
% Please add line breaks to long equations when possible in order to fit our 2-column layout. 
%
% For inline equations, please do not include punctuation within the math environment unless this is part of the equation.
%
% For spaces within the math environment please use the \; or \: commands, even within \text{} (do not use smaller spacing as this does not convert well).
%
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{bm}
\usepackage{hyperref}

% line numbers
\usepackage{lineno}

% ligatures disabled
\usepackage{microtype}
% \DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
%\usepackage{rotating}

% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}

% Use doublespacing - comment out for single spacing
% \usepackage{setspace} 
% \doublespacing

\usepackage{graphicx}
\usepackage{subfigure}

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}

@@

\pagestyle{myheadings}

%% Include all macros below. Please limit the use of macros.

@@

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{A simple model-based approach to inferring and visualizing cancer mutation signatures}
}
% Insert Author names, affiliations and corresponding author email.
\\
Yuichi Shiraishi$^{1,\ast}$, 
Georg Tremmel$^{1}$, 
Satoru Miyano$^{1}$, 
Matthew Stephens$^{2,3, \ast}$
\\
\bf{1} Laboratory of DNA Information Analysis, Human Genome Center, Institute of Medical Science, The University of Tokyo, Tokyo, Japan
\\
\bf{2} Dept. of Human Genetics, University of Chicago, Chicago, Illinois, United States of America
\\
\bf{3} Dept. of Statistics, University of Chicago, Chicago, Illinois, United States of America
% \bf{3} Author3 Dept/Program/Center, Institution Name, City, State, Country
\\
$\ast$ E-mail: yshira@hgc.jp; mstephens@uchicago.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

@@

Here we present new methods for modelling, identifying and visualizing such mutation signatures. Our methods
greatly simplify mutation signature models compared with existing approaches, reducing the number of parameters by orders of magnitude even while increasing the contextual factors (e.g.~number of flanking bases) that are accounted for. This improves both sensitivity and robustness of inferred signatures. We also provide a new intuitive way to visualize the signatures, analogous to the use of sequence logos to visualize transcription factor binding sites.

@@

%This paragraph is nice, but could be removed if short on words?
Our work draws on ideas from several tools that are widely used in other contexts. The model for each signature is connected with position weight matrix models for transcription factor binding, and the overall framework is based on probabilistic models that are closely
connected with ``mixed-membership models'' which are widely used in population genetic admixture analysis, and in machine learning for document clustering. We argue that recognizing these relationships should help improve
understanding of mutation signature extraction problems,
and suggests ways to further improve the statistical methods.

@@

% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}

@@

@@

Figure \ref{mutSig_example} illustrates the way that our new representation of signatures can simply capture
a previously-identified signature \cite{pmid25362482, pmid23318258} and provides
an easily-interpretable visualization of the signature 
that is analogous to  sequencing logos \cite{pmid2172928}.
We particularly note how the key elements of this mutation signature are more immediately visually apparent than with visualizations of the full vector of probabilities used by existing approaches.

@@

Suppose each somatic mutation has $L$ mutation features, $\bm{m} = (m_1, m_2, \cdots, m_L)$, 
where each $m_l$ can take $M_l$ discrete values. Also, let $\bm{M} := (M_1, \cdots, M_L)$.
For example, when taking account of 6 substitution patterns and $\pm 2$ flanking sites,
$\bm{M} = (6, 4, 4, 4, 4)$.
See Supplementary Table 1 for other examples.

@@

This generative model is summarized in Figure \ref{methodOverview}.
The key parameters in this model
are the membership proportions for each sample, $\bm{q}_i$, and the mutation signature parameters, $\bm{F}_k$. 
We estimate these parameters by maximizing likelihood using an EM algorithm. 
A simulation study demonstrates that the estimation method can reproduce the mutation signature very accurately 
provided enough mutations and samples are available (see Supplementary Material and Supplementary Figure 1). 
See Methods for more details, how to select $K$, etc.

@@

We consider a mutation pattern to consist of the substitution pattern, the $\pm 2$ flanking bases, and the transcription strand direction. Thus each signature is characterized by 18 parameters in our independent model, and by 3071 parameters in the full model. After analyzing the data with various number of mutation signatures $K$,
we selected $K=3$ signatures for these analyses (Supplementary Material, Supplementary Figure 2, 3 and 4).

The inferred APOBEC signature under the independent model
shows a clear depletion of G base at the $-2$ position,
which is consistent to the previous study \cite{pmid23318258} and results in the next subsection.
In contrast, for the full model, this tendency is rather mild (Figure \ref{UCUT:APOBEC_full5_sig}, \ref{UCUT:AA_full5_sig} and Supplementary Figure 5).
The inferred AA mutation signature has no clear characteristics at the $-2$ position.
These results suggest that our independent model has the potential
to identify signatures in more detail and with less data than existing approaches based on the full model.

@@

The results (Figures \ref{UCUT:APOBEC_downsampling} and \ref{UCUT:AA_downsampling})
confirm that the results of the independent model 
are substantially more robust to reductions in data size than the full model. Indeed, mutation signatures inferred using the independent model with only 10\% of the data 
remain highly similar to the signatures inferred from the full data; by comparison the full model shows a much larger drop-off in similarity, especially in the APOBEC signature where even using 50\% of the data gives a substantial drop-off in similarity. Both methods found the AA signature 
easier to recover than the APOBEC signature.
We believe that this is because the number of T $>$ A substitutions at GpTpC sites are far more frequent in this dataset.

@@

To provide a more comprehensive practical illustration of
our method, we applied it to somatic mutation data from 30 cancer types \cite{pmid23945592}.
We applied the method to each cancer type separately to assess similarity of estimated signatures across cancer types.
For each cancer type we selected the number of signatures $K$ by fitting the model with increasing $K$ and examining the log-likelihood, bootstrap errors, and correlation of membership parameters.The selected values of $K$ are given in 
Supplementary Table 2.
Also, we simply removed somatic mutations located in an intergenic region to include transcription strand biases as mutation features.
Finally, we merged similar mutation signatures across different cancer types 
(when their Frobenius Distance were $<$ $0.6$). 
% XXX WHAT KIND OF CLUSTERING?
 % (see Supplementary Material for the values of the log-likelihoods and bootstrap errors for each cancer type).

Figures \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarise the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

Signatures 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
appear likely to be associated with deregulated activity of the error-prone polymerase Pol $\epsilon$. 
In previous analyses of these data \cite{pmid23945592}, 
the signature for Pol $\epsilon$ dysfunction was represented by a single signature (their ``signature 10'').
In contrast our new approach uses two signatures.
Since these signatures are highly correlated,
and appear connected by
a single biological mechansim, we certainly do not argue
that inferring them as a single signature is ``wrong". However, splitting them into two signatures does help highlight certain features. Specifically,
signature 1 shows a transcription strand bias whereas signature 8 does not, and this is true for both colorectal and uterus cancers (Supplementary Figures 6(c, d)).
This strand bias may be connected with the enrichment of 
C $>$A at TpCpT mutations in leading strands of replication forks
observed by \cite{pmid25228659}. Although replication strand bias is different from transcription strand bias, these two biases may be connected through the fact that replication origins prefer transcription start sites \cite{pmid23187890}.

These signatures also illustrate the ability of our model
to help highlight sequence context effects beyond the immediate flanking bases. Specifically, both signatures 1 and 8  show an elevated frequency of the T base at position $-2$,
and signature 1 also shows slightly elevated frequency of the T base at position $+2$
 (Figures \ref{POLE1_two5prime} and \ref{POLE2_two5prime}, Supplementary Figure 6(c, d)).
A previous study of Pol $\epsilon$  \cite{pmid25228659} 
found that a nonsense mutation R23X of TP53 is enriched in cancers with Pol $\epsilon$ defects.
In fact, the pattern of this mutation is C $>$ T at TpTpCpGpA, closely matching signature 8.
This illustrates that the inclusion of $\pm 2$ bases 
into signatures may be helpful for identifying underlying mechanisms.

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (Supplementary Figure 7 (a)). Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (Supplementary Figure 7 (b,c)), and signature 2 disappeared when we removed this sample (Supplementary Figure 8).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

%XX Also important, making a connection, but not the new method XX
Signature 4 (C $>$ A at CpCpG) observed in kidney clear cell carcinomas, lung adenocarcinomas and melanomas 
seems to correspond to the ``signature R2''  detected in the same cancer types (plus lung squamous carcinomas) in 
\cite{pmid23318258} (see their Supplementary Figures).
Again our analysis highlights additional contextual information, with a strikingly elevated frequency of base C at the $-2$ position (Supplementary Figure 9 (a,b,c)).
However, for each cancer type, only a few samples support this signature (see Supplementary Figure 9 (d, e, f)),
and the corresponding signature 
could not be validated in the previous study: most somatic mutations corresponding to that signature could not be validated by re-sequencing or visual inspection of BAM files using genomic viewers. Again, further investigation
yielded a potential explanation for this finding:
this signature largely matches that of a putative artifact caused by oxidation of DNA during acoustic shearing \cite{pmid23303777}, 
and we conclude that this signature, and the corresponding signature in previous work, are likely artefactual.
Although not of direct biological interest,
identifying artefactual signatures could be helpful in removing false positive mutations.

>>>>

% Template for PLoS
% Version 3.1 February 2015
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% added by yshira( for using subfigure... this will be not necessary at the time of submission)
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{bm}
\usepackage{comment}

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

@@

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

@@

\begin{document}
\vspace*{0.35in}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{A simple model-based approach to inferring and visualizing cancer mutation signatures}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Yuichi Shiraishi\textsuperscript{1, *},
Georg Tremmel\textsuperscript{1},
Satoru Miyano\textsuperscript{1},
Matthew Stephens\textsuperscript{2,3,*},
\\
\bigskip
\bf{1} Laboratory of DNA Information Analysis, Human Genome Center, Institute of Medical Science, The University of Tokyo, Tokyo, Japan
\\
\bf{2} Dept. of Human Genetics, University of Chicago, Chicago, Illinois, United States of America
\\
\bf{3} Dept. of Statistics, University of Chicago, Chicago, Illinois, United States of America
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
% \Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
% \ddag These authors also contributed equally to this work.

% Current address notes
% \textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
% \dag Deceased

% Group/Consortium Author Note
% \textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* yshira@hgc.jp; mstephens@uchicago.edu

\end{flushleft}

% Please keep the abstract below 300 words
\section*{Abstract}

@@

Here we present new methods for modelling, identifying and visualizing such mutation signatures. Our methods
greatly simplify mutation signature models compared with existing approaches, reducing the number of parameters by orders of magnitude even while increasing the contextual factors (e.g. the number of flanking bases) that are accounted for. This improves both sensitivity and robustness of inferred signatures. We also provide a new intuitive way to visualize the signatures, analogous to the use of sequence logos to visualize transcription factor binding sites.

@@

%This paragraph is nice, but could be removed if short on words?
% Our work draws on ideas from several tools that are widely used in other contexts. The model for each signature is connected with position weight matrix models for transcription factor binding, and the overall framework is based on probabilistic models that are closely
The overall framework of our work is based on probabilistic models that are closely
connected with ``mixed-membership models'' which are widely used in population genetic admixture analysis, and in machine learning for document clustering. We argue that recognizing these relationships should help improve
understanding of mutation signature extraction problems,
and suggests ways to further improve the statistical methods.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}

@@

\linenumbers

@@

Fig. 1 illustrates the way that our new representation of signatures can simply capture
a previously-identified signature \cite{pmid25362482, pmid23318258} and provides
an easily-interpretable visualization of the signature 
that is analogous to  sequencing logos \cite{pmid2172928}.
We particularly note how the key elements of this mutation signature are more immediately visually apparent than with visualizations of the full vector of probabilities used by existing approaches.

\begin{figure}[h]

\caption{{\bf Examples of visualizations and parameter values for the mutation signatures of the unconstrained (full) model and our independent model, 
where substitution patterns, two 5' and 3' bases and transcription strand direction are considered as mutation features.}
(A) The barplots are divided by 6 substitution patterns and transcription strand direction.
In each division, 256 bars show joint probabilities of up to two base 5' and 3' bases with substitution patterns and transcription strands
(ApApNpApA, ApApNpApC, ApApNpApG, ApApNpApT, $\cdots$, TpTpNpTpT).
(B, C) An example mutation signature representation and parameter values from our independent model, where mutation features (substitution patterns, two 5' and 3' bases and strand direction) 
are assumed to be independent ($L=6$, $\bm{M} = (6, 4, 4, 4, 4, 2)$).
In the bottom five rectangles, the width of each box represents the frequencies of bases (A, C, G and T) at the substitution and flanking site. 
To highlight the most informative flanking sites, the heights of flanking site boxes are scaled by ($1 + 0.5 \times \log \sum_{n = A,C,G,T} f_n^2$, where $f_n$ is the parameter for each base), which can be interpreted as ($1 - 0.5 \times \text{R\'enyi entropy}$) \cite{rrnyi1961measures}. This is analogous to the information content scaling used in sequencing logos.
In the top rectangle, the height of each box represents the conditional frequencies of mutated bases for each original base (C and T).
In the upper right, the height of the $+$ box represents the frequencies of mutations in the coding strand (or the plus strand, the sense strand and the untranscribed strand)
whose nucleotide sequences directly corresponds to mRNA, 
whereas the height of $-$ box represents those in the template strand (or the minus strand, the antisense strand, the transcribed strand and the noncoding strand)
whose sequences are copied during the synthesis of mRNA.
}
\label{mutSig_example}

\end{figure}

@@

Suppose each somatic mutation has $L$ mutation features, $\bm{m} = (m_1, m_2, \cdots, m_L)$, 
where each $m_l$ can take $M_l$ discrete values. Also, let $\bm{M} := (M_1, \cdots, M_L)$.
For example, when taking account of 6 substitution patterns and $\pm 2$ flanking sites,
$\bm{M} = (6, 4, 4, 4, 4)$.
See S1 Table for other examples.

@@

This generative model is summarized in Fig. \ref{methodOverview}.
The key parameters in this model
are the membership proportions for each sample, $\bm{q}_i$, and the mutation signature parameters, $\bm{F}_k$. 
We estimate these parameters by maximizing likelihood using an EM algorithm. 
A simulation study demonstrates that the estimation method can reproduce the mutation signature very accurately 
provided enough mutations and samples are available (see S1 Text). 
See Methods for more details, how to select $K$, etc.

\begin{figure}[h]
% \includegraphics[width=450pt]{methodOverview.pdf}
\caption{{\bf An overview of the generative model of somatic mutations proposed in this paper.} 
Suppose there are three types of mutation sources (mutation signatures) such as ultraviolet, tobacco smoking chemicals and transcription coupled repairs.
Each cancer genome has ratios showing which types of mutation sources are contributing to its mutations (membership parameters).
The generative model of the pattern of each mutation is:
first, one of the mutation signatures is chosen according to the membership parameter.
Second, according to the selected mutation signature, mutation features such as substitution pattens and flanking bases are generated 
by the corresponding multinomial distributions for that mutation signature.
}
\label{methodOverview}
\end{figure}

@@

We consider a mutation pattern to consist of the substitution pattern, the $\pm 2$ flanking bases, and the transcription strand direction. Thus each signature is characterized by 18 parameters in our independent model, and by 3071 parameters in the full model. After analyzing the data with various number of mutation signatures $K$,
we selected $K=3$ signatures for these analyses (see S2 Text).

The inferred APOBEC signature under the independent model
shows a clear depletion of G base at the $-2$ position,
which is consistent to the previous study \cite{pmid23318258} and results in the next subsection ((Figs. 3A and 3B)).
In contrast, for the full model, this tendency is rather mild (Figs. 3C, 3D and S1).
The inferred AA mutation signature has no clear characteristics at the $-2$ position.
These results suggest that our independent model has the potential
to identify signatures in more detail and with less data than existing approaches based on the full model.

\begin{figure}[h]
\caption{{\bf The mutation signatures for the UCUT data, and the results of downsampling experiments.}
3072 elements in the full model mutation signatures were shown divided by 6 substitution patterns and strand directions.
(A) APOBEC signature for the independent model.
(B) AA signature for the independent model.
(C) APOBEC signature for the full model.
(D) AA signature for the full model.
(E) APOBEC signature stability.
(F) AA signature stability.}
\label{UCUT}
\end{figure}

@@

The results (Figs. 3E and 3F)
confirm that the results of the independent model 
are substantially more robust to reductions in data size than the full model. Indeed, mutation signatures inferred using the independent model with only 10\% of the data 
remain highly similar to the signatures inferred from the full data; by comparison the full model shows a much larger drop-off in similarity, especially in the APOBEC signature where even using 50\% of the data gives a substantial drop-off in similarity. Both methods found the AA signature 
easier to recover than the APOBEC signature.
We believe that this is because the number of T $>$ A substitutions at GpTpC sites are far more frequent in this dataset.

@@

To provide a more comprehensive practical illustration of
our method, we applied it to somatic mutation data from 30 cancer types \cite{pmid23945592}.
We applied the method to each cancer type separately to assess similarity of estimated signatures across cancer types.
For each cancer type we selected the number of signatures $K$ by fitting the model with increasing $K$ and examining the log-likelihood, bootstrap errors, and correlation of membership parameters.The selected values of $K$ are given in S2 Table.
Also, we simply removed somatic mutations located in an intergenic region to include transcription strand biases as mutation features.
Finally, we merged similar mutation signatures across different cancer types 
(when their Frobenius Distance were $<$ $0.6$). 
% XXX WHAT KIND OF CLUSTERING?
 % (see Supplementary Material for the values of the log-likelihoods and bootstrap errors for each cancer type).

Figs. \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarise the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

\begin{figure}[h]
\centering
% \includegraphics[width=16cm,height=17.5cm]{AlexandrovEtAl_mergedSignature.eps}
\caption{{\bf The summary of mutation signatures obtained in a reanalysis of the data of the previous study \cite{pmid23945592} using the proposed method.}
Here, the substitution patterns and two 5' and 3' bases from the mutated sites are taken into account as mutation features.
First, mutation signatures were estimated separately in each cancer type, and then similar signatures were merged (see text).
}
\label{nature2013_sig_summary}
\end{figure}

\begin{figure}[h]
\centering
% \includegraphics[width=15cm,height=15cm]{corrplot.eps}
\caption{{\bf The summary of membership of each mutation signature across 30 cancer types obtained using the proposed method.}}
\label{nature2013_sig_member}
\end{figure}

Signatures 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
appear likely to be associated with deregulated activity of the error-prone polymerase Pol $\epsilon$. 
In previous analyses of these data \cite{pmid23945592}, 
the signature for Pol $\epsilon$ dysfunction was represented by a single signature (their ``signature 10'').
In contrast our new approach uses two signatures.
Since these signatures are highly correlated,
and appear connected by
a single biological mechansim, we certainly do not argue
that inferring them as a single signature is ``wrong". However, splitting them into two signatures does help highlight certain features. Specifically,
signature 1 shows a transcription strand bias whereas signature 8 does not, and this is true for both colorectal and uterus cancers (Figs. S2C S2D).
This strand bias may be connected with the enrichment of 
C $>$A at TpCpT mutations in leading strands of replication forks
observed by \cite{pmid25228659}. Although replication strand bias is different from transcription strand bias, these two biases may be connected through the fact that replication origins prefer transcription start sites \cite{pmid23187890}.

These signatures also illustrate the ability of our model
to help highlight sequence context effects beyond the immediate flanking bases. Specifically, both signatures 1 and 8  show an elevated frequency of the T base at position $-2$,
and signature 1 also shows slightly elevated frequency of the T base at position $+2$
 (Figs. 6B, 6C ,S6C and S6D).
A previous study of Pol $\epsilon$  \cite{pmid25228659} 
found that a nonsense mutation R23X of TP53 is enriched in cancers with Pol $\epsilon$ defects.
In fact, the pattern of this mutation is C $>$ T at TpTpCpGpA, closely matching signature 8.
This illustrates that the inclusion of $\pm 2$ bases 
into signatures may be helpful for identifying underlying mechanisms.

\begin{figure}[h]
% \subfigure[APOBEC signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=14.5cm,height=6cm]{APOBEC_two5prime.eps}
%    \label{APOBEC_two5prime}}
% \subfigure[POLE1 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE1_two5prime.eps}
%   \label{POLE1_two5prime}}
% \quad
% \subfigure[POLE2 signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{POLE2_two5prime.eps}
%   \label{POLE2_two5prime}}
% \subfigure[UV signature intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{UV_two5prime.eps}
%   \label{UV_two5prime}}
% \quad
% \subfigure[The signature 11 intensities at two 5' to the mutated site]{%
%   \includegraphics[width=4.5cm,height=6cm]{LMST_two5prime.eps}
%   \label{LMST_two5prime}}
\caption{{\bf The estimated frequencies of bases at two 5' to the mutated site for each cancer type.}
The bar heights show the estimated frequency for bases A, C, G and T at two 5' to the mutated site.
The error bars show bootstrapped standard errors.
(A) The signature 13 (APOBEC signature)  intensities at two 5' to the mutated site.
(B) The signature 1 (the first POL $\epsilon$ signature) intensities at two 5' to the mutated site.
(C) The signature 8 (the second POL $\epsilon$ signature) intensities at two 5' to the mutated site.
(D) The signature 10 (ultraviolet signature) intensities at two 5' to the mutated site.
(E) The signature 11 intensities at two 5' to the mutated site.
}
\label{two5prime}
\end{figure}

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (``signature 14'' \cite{pmid23318258}). Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (see Figs. S3A and S3B), and signature 2 disappeared when we removed this sample (see Fig. S4).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

%XX Also important, making a connection, but not the new method XX
Signature 4 (C $>$ A at CpCpG) observed in kidney clear cell carcinomas, lung adenocarcinomas and melanomas 
seems to correspond to the ``signature R2''  detected in the same cancer types (plus lung squamous carcinomas) in 
\cite{pmid23318258} (see their Supplementary Figures).
Again our analysis highlights additional contextual information, with a strikingly elevated frequency of base C at the $-2$ position (Figs. S5A, S5B and S5C).
However, for each cancer type, only a few samples support this signature (see Figs. S5D, S5E and S5F),
and the corresponding signature 
could not be validated in the previous study: most somatic mutations corresponding to that signature could not be validated by re-sequencing or visual inspection of BAM files using genomic viewers. Again, further investigation
yielded a potential explanation for this finding:
this signature largely matches that of a putative artifact caused by oxidation of DNA during acoustic shearing \cite{pmid23303777}, 
and we conclude that this signature, and the corresponding signature in previous work, are likely artefactual.
Although not of direct biological interest,
identifying artefactual signatures could be helpful in removing false positive mutations.
